Standartox: Standardizing toxicity data

### 2020-04-07 article/article.tex
update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data:
\dataset{https://cfpub.epa.gov/ecotox/}
%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately. If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal. In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}

@@

We therefore developed Standartox, a tool and database that aims to overcome the limitations of other databases by continuously incorporating the ever-growing number of test results in an automated process workflow that ultimately leads to a harmonized ecotoxicity data collection and provides methods to derive single aggregated ecotoxicity values for a specific chemical-organism test combination. Standartox makes use of the publicly available and quarterly updated ECOTOX database \citep{usepa_ecotox_2019} and restricts the data to commonly used endpoints in ecotoxicology, such as half maximal effective concentration (EC\textsubscript{50}) or no-observed-adverse-effect concentrations (NOEC), leading to about 600,000 ecotoxicological test results including 8000 chemicals, tested on about 8000 taxa. Standartox users can filter test results according to several parameters, e.g. refining a search for ecotoxicity data on organisms occurring in specific habitats or regions of the world. Above all, Standartox aggregates ecotoxicological test results in a standardized way, by calculating the minimum, the geometric mean and the maximum of the results for each chemical and the associated, user-defined test parameters. Hence, this reduces the variability between risk assessments that are due to the selection of different ecotoxicological test data \citep{mark_analysis_1998}. Thereby, Standartox provides the basis for reproducible science and combines information from different sources to simplify the derivation of risk indicators such as Species Sensitivity Distributions (SSD) and Toxic Units (TU), which represent two prominent concepts to assess effects on organisms in ecotoxicology \citep{posthuma_species_2002, kefford_definition_2011, schafer_effects_2011}. Besides aggregating ecotoxicological test results, Standartox provides a concise overview of the tested chemicals, allowing the identification of potential knowledge gaps. Moreover, Standartox could help in reducing the millions of animals used for toxicity testing each year by facilitating access to ecotoxicity data \citep{hartung_chemical_2009}. Standartox comes with two front-ends, a web application: \url{standartox.uni-landau.de} and an R \citep{rcoreteam_language_2017} package \textit{standartox}, providing convenience structures and thereby largely reducing processing time for users.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data Description - required}
Standartox constitutes a collection of quality checked ecotoxicological test results. It is build on the ECOTOX database \citep{usepa_ecotox_2019} whose data are processed, cleaned and harmonized to retrieve comparable toxicity endpoints. Subsequently, filter and aggregation methods are created to allow for the retrieval of single toxicity equivalents for specific experimental conditions. The ECOTOX database is updated quarterly, providing on average 5228 (2014 - 2019) new toxicity entries. These are included in Standartox with each update.

@@

\subsection{Perspectives}
Novel predictive frameworks incorporating chemical mode of action and species traits emphasize the need for holistic and automated analyses of large-scale ecotoxicological data \citep{malaj_evolutionary_2016, vandenberg_modeling_2019}. Indeed, the increasing amount of data from ecotoxicological tests and experiments that is becoming available has elicited several initiatives to harmonize these data. These initiatives partly aim for overlapping goals, yet have limitations or objectives that distinguish them from Standartox:

\textbf{Comptox}, is a web tool published by the EPA which, similar to Standartox allows for filtering test results, the retrieval of additional chemical information and the estimation of toxicity data, such as 48 h \textit{Daphnia magna} LC\textsubscript{50} values. However, toxicity estimations are limited to standard test organisms (e.g. \textit{Daphnia magna}), and the tool lacks the possibility for automated retrieval \citep{williams_comptox_2017}.

The \textbf{EnviroTox} database which also uses, amongst others the ECOTOX database as an input has recently been published \citep{healthandenvironmentalsciencesinstitutehesi_envirotox_2019, connors_creation_2019}. In contrast to Standartox, EnviroTox  is restricted to selected aquatic organisms (i.e. fish, amphibians, invertebrates and algae) and experimental durations (at least 24 h) and uses a rule-based algorithm to derive single ecotoxicity values. Besides, EnviroTox provides additional information on toxicity endpoints, such as acute or chronic classifications and mode of action assignments. We intentionally omitted such classifications given that the approach to classification may vary with the purpose of the study. The EnviroTox database allows for an aggregation into single toxicity values for individual taxa, whereas Standartox performs this aggregation for individual taxa-chemical combinations. However, the Standartox results for individual taxa-chemical combinations could easily be aggregated across chemicals in a second step to provide a similar aggregation as that performed in EnviroTox.

The \textbf{Etox} database, like the ECOTOX database collects and provides filter methods for ecotoxicological test information.

The \textbf{PPDB} provides data only on pesticides and as mentioned before provides single quality controlled values only for commonly used taxa, e.g. \textit{D. magna} or \textit{R. subcapitata}.

\begin{table}[H]
    \caption{Overview on databases that provide ecotoxicological data. Abbreviations: \textbf{ALL:} Most important test parameters, including chemical, taxon, duration for filtering ecotoxicological data are incorporated. \textbf{WEB:} Accessible via a web application through a graphical user interface. \textbf{API:} Accessible via an application programming interface.}
    \label{tab:database-differences}
    \centering
    \begin{tabular}{m{3cm}m{3cm}m{2cm}m{2cm}m{1cm}}
    \toprule
    \textbf{Database} & \textbf{Publisher} & \textbf{Filter} & \textbf{Aggregation, Selection} & \textbf{Access} \\
    \midrule
    Comptox & Environmental Protection Agency & Chemical & no & WEB, file \\
    Ecotox & Environmental Protection Agency & ALL & no & WEB, file \\
    EnviroTox & Health and Environmental Sciences Institute & ALL & chemical, organism \\
    Etox & Umweltbundesamt & ALL & no & WEB \\
    Pesticides Properties DataBase (PPDB) & University of Hertfordshire & fixed values & manual selection & WEB, file \\
    Standartox & this article & ALL & chemical, organism & API, WEB \\
    \bottomrule
\end{tabular}
\end{table}

In summary, none of the above mentioned initiatives aim for an automated and standardized aggregation method of exposure endpoints for individual chemicals. In addition, they lack the possibility to access the databases through a common high level programming language, such as R. As outlined above, toxicity estimates from different studies can vary strongly due to a wide range of experimental conditions such as pH, temperature and conductivity \citep{rosenkrantz_influence_2013, li_temperature_2011}. Integrating these conditions into the aggregated estimates would certainly improve toxicity estimates. However, the current implementation of Standartox omits these conditions, because the ECOTOX database only provides sparse records on experimental conditions. The most frequently provided experimental conditions are temperature (77~\%), pH (56~\%), hardness (27~\%), dissolved oxygen (18~\%), Alkalinity (15~\%) and salinity (9~\%). For all other conditions less than 5~\% of data entries are available. A text-mining approach, where a literature reference is associated with ecotoxicity raw data, iterating through the individual publications could potentially increase this number. E.g. \citet{compson_linking_2018} successfully applied text-mining techniques to retrieve species trait data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
An automated processing pipeline downloads the quarterly released ECOTOX database, performs several preparation steps on it and exports a final Standartox data set. This data set is accessible via a web application and an application programming interface (API). An API provides the means for machine communication between a host and a client and thus allows scriptable data queries. To facilitate the API access, the R \citep{rcoreteam_language_2017} package \textit{standartox} is built. All data presented in this paper are derived from the Standartox build based on the ECOTOX release from the 12.12.2019. The code for Standartox is located in two Github repositories \url{https://github.com/andschar/standartox-build} and \url{https://github.com/andschar/standartox}. The former contains code to process the data and to build the web application and the API, the latter contains code to build the R-package. Most of the code is written in R 3.6.1 and associated packages (List: \ref{lis:r-packages}) and SQL for PostgreSQL 9.6.1.

\subsection{Processing}
Standartox downloads the quarterly released ECOTOX database and builds it into a local PostgreSQL database \citep{szocs_build_2019}. Subsequently, Structured Query Language (SQL) functions for further processing the data are implemented. In addition lookup tables that enable the conversion of units such as duration and concentration are created. A meta-table providing information such as the release version of the ECOTOX database is added. Then, provided Chemical Abstracts Service Numbers (CAS) and taxonomic names are used to query additional information from publicly available databases on chemicals and organisms, respectively. This includes the Compendium of Pesticide Common Names \citep{wood_compendium_2019}, the Chemical Entities of Biological Interest (ChEBI) database \citep{hastings_chebi_2016}, the Chemical Identifier Resolver (CIR) service \citep{nationalinstitutesofhealthnih_chemical_2019}, the Pubchem database \citep{kim_pubchem_2016}, Eurostat \citep{europeancommission_eurostat_2019} and Wikidata \citep{vrandecic_wikidata_2014} for chemicals and the World Register of Marine Species (WoRMS) \citep{wormseditorialboard_world_2018} and the Global Biodiversity Information Facility (GBIF) \citep{gbif_gbif_2019} (Table~\ref{tab:data-base-additional}) for habitat and spatial distribution of organisms. Given that CAS and taxonomic names can be ambiguous, e.g. the genus \textit{Eisenia} can refer to an algae and a worm, we first match them against the database specific identifiers and subsequently use them to retrieve the actual data. In a next step the data is added to Standartox to enable filtering for specific chemical roles (e.g. drug, metal, pesticide, personal care product) and classes (e.g. pyrethroid, carbamate) as well as spatial distribution (i.e. continents) and habitat preferences (e.g. freshwater) of individual taxa. Taxa that were not identified to at least genus level are excluded, because relative toxicity comparisons have been shown to be not meaningful for higher taxonomic levels \citep{rainbow_trace_2002, buchwalter_differences_2005, malaj_physiological_2012}. Finally, the Standartox data set is compiled, which includes the harmonisation of data, e.g. through conversion of test concentration and duration units. 1237 distinct concentration units are converted to X harmonized ones, as are the 126 distinct duration units to X. Standartox retains only duration units that can unambiguously be converted to hours. To guarantee appropriate unit conversion and harmonisation, we compared an automated to a manual conversion for each of the distinct concentration and duration units. Furthermore, the units are cleaned, for example through removing additional information in the field such as \textit{food}, \textit{soil}, \textit{ai} that are also coded in other variables and hinder the processing of units. Concentrations that are given as rates such as per day (e.g. mg/kg/day) are multiplied by the days of the test and then converted. Experimental endpoints are restricted to three groups, namely NOEX, LOEX and XX50. Other endpoints, such as Bioconcentration factors, non-half maximal effective concentrations (e.g. IC10, EC25, LD99) or maximum acceptable toxicant (MATC) concentrations are removed. Along with that, a catalog, listing all distinct entries and value ranges, for categorical and continuous variables, respectively, is created. Finally, we run quality control scripts that check the accuracy of the data. The compiled Standartox data set together with the catalog is exported and accessible via the web application and the API, through the R package.

\begin{figure}[H]
    \centering
    \includestandalone[scale=0.75]{article/tikz/standartox-organigram}
    \caption{Organigram of Standartox. The ECOTOX knowledge base is downloaded quarterly and processed (i.e. query additional information with CAS and taxa name and conversion of concentration and duration units. Subsequently, a Standartox data set is compiled together with filter and aggregation methods. Thus, users can access the Standartox data set and filter and aggregate through a web application and an R-package.}
    \label{fig:stx-organigram}
\end{figure}

\subsection{Application methods}
When accessed, the web application and the API load the compressed serialized Standartox data into memory and allow the user to interact with it. The user can then call the functions stx\_filter() and stx\_aggregate() that filter and aggregate the data according to specific parameters (Table~\ref{tab:app-parameters}). The interactive web application is built in R using the shiny framework, which runs with the help of a shiny server \citep{chang_shiny_2018} on \url{standartox.uni-landau.de}. The API is built by using the R package \textit{plumber} \citep{trestletechnologyllc_plumber_2018}, which allows for the creation of Representational State Transfer (REST) APIs from R. REST is a software architectural style that defines web service communication rules. The API is reachable via the URL \url{http://139.14.20.252} and port 8000. Four API-endpoints (\textit{/catalog}, \textit{/filter}, \textit{/aggregate} and \textit{/meta}) can be queried (Table~\ref{tab:api-endpoints}). The \textit{/catalog} API-endpoint returns a JavaScript Object Notation (JSON) file containing a catalog of possible filter parameters to choose from. The \textit{/filter} returns the filtered Standartox table as a compressed serialized binary file created by the R \textit{fst} package \citep{klik_fst_2019}, to reduce size and allow for fast user queries. The \textit{/aggregate} API-endpoint returns an R function allowing the client to aggregate the filtered data locally. Lastly, the \textit{/meta} API-endpoint returns a JSON file with meta information, such as the timestamp of the request and the used Standartox version. The API is designed to be used with the R-package \textit{standartox} and therefore uses serialization methods specific to R (rds() from the R package \textit{base} and fst() from the package \textit{fst}). To facilitate the API usage the R-package \textit{standartox} is created.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{User Notes - optional}
Users can access Standartox either via the web application \url{standartox.uni-landau.de} or via the R-package \textit{standartox}. By accessing the web application, users can filter and download the resulting data sets as a comma-separated values (csv) file. Users of the R-package can directly load the data within R. The R-package provides the two functions \textit{stx\_catalog()} and \textit{stx\_query()}. The first command queries a catalog of possible Standartox parameters into an R list object. The latter allows users to set the Standartox filter parameters and to fetch the actual data. It returns an R list of three tables (i.e. R data.frames) containing the filtered data set, the aggregated data set and a table with the meta information retrieved from the API endpoints. A short R-code example is given below (Listing \ref{lis:standartox-example}) and a detailed description on the usage of the R package is provided on its Github page: \url{https://github.com/andschar/standartox}.

\begin{lstlisting}[
    caption = {Sample code to access the Standartox data base through the API with the help of the R-package. \textit{stx\_catalog()} returns a catalog of possible filter and aggregation parameters. \textit{stx\_query()} returns the Standartox object, a list of the filtered and the aggregated data as well as a meta data entry.},
    label = {lis:standartox-example}]

# install
install.packages('remotes')
remotes::install_github('andschar/standartox') # package not yet on CRAN
# retrieve catalog    
require(standartox)
catal = stx_catalog()
# retrieve data
l = stx_query(cas = '1071-83-6',
                endpoint = 'XX50',
                taxa = 'Oncorhynchus',
                duration = c(24, 120))
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion - optional}
% This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
Due to the steady incorporation of new ecotoxicity data, the aggregated values produced by Standartox can be subject to change with future updates. We regard this as an advantage rather than a drawback because other published works that aim in a similar direction often constitute a singular effort or require manual work for each update. Standartox, in contrast automates the update process, yet still provides access to its older versions, assuring reproducability and version control. In comparison to rule-based approaches for the derivation of single ecotoxicity values, Standartox has the advantage to be free from the subjectivity of a set of human-induced rules. Above all, Standartox provides quick access through its design to be queried via the R language. Due to an increased amount of available ecotoxicological test data, it becomes fundamental to provide and distribute ecotoxicity information in adequate formats, both easily accessible for humans and easily processable for machines. Standartox meets these requirements and puts its focus on the aggregation of toxicity data, thereby adding a piece to the puzzle of modern ecotoxicological data analyses.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{conceptualization and methods, A.S.,R.B.S. and V.C.S.; software development and validation, A.S.; data curation and analysis, A.S.; writing--original draft preparation, A.S.; writing--review and editing, A.S., V.C.S. and R.B.S.; supervision, R.B.S.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections. This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{The authors declare no conflict of interest.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{The following abbreviations are used in this manuscript:\\
\noindent 
\begin{tabular}{@{}ll}
API & Application programming interface \\
CAS & Chemical abstracts service registry Number \\
ChEBI & Chemical Entities of Biological Interest database \\
CIR & Chemical Identifier Resolver service \\
E/LC/D\textsubscript{50} &  Half maximal effective/lethal concentration/dose \\
ECOTOX & US EPA Ecotox Knowledgebase \\
GBIF & Global Biodiversity Information Facility \\
JSON & JavaScript Object Notation file format \\
LOEC/L & Lowest observed effect levels/concentration \\
NOEC/L & No observed effect levels/concentration \\
PPDB & Pesticides Properties DataBase \\
QSAR & Quantitative structure–activity relationship \\
REST & Representational state transfer software architectural style \\
SSD & Species sensitivity distribution \\
TU & Toxic unit \\
WoRMS & World Register of Marine Species \\
\end{tabular}
}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data:
%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately. If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal. In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}

@@

We therefore developed Standartox, a tool and database that aims to overcome the limitations of other databases by continuously incorporating the ever-growing number of test results in an automated process workflow that ultimately leads to a harmonized ecotoxicity data collection and provides methods to derive single aggregated ecotoxicity values for a specific chemical-organism test combination. Standartox makes use of the publicly available and quarterly updated ECOTOX database \citep{usepa_ecotox_2019} and restricts the data to commonly used endpoints in ecotoxicology, such as half maximal effective concentration (EC\textsubscript{50}) or no-observed-adverse-effect concentrations (NOEC), leading to about 600,000 ecotoxicological test results including about 8000 chemicals, tested on about 10000 taxa in the current version. Standartox users can filter test results according to several parameters, e.g. refining a search for ecotoxicity data on organisms occurring in specific habitats or regions of the world. Above all, Standartox aggregates ecotoxicological test results in a standardized way, by calculating the minimum, the geometric mean and the maximum of the results for each chemical and the associated, user-defined test parameters. Hence, this reduces the variability between risk assessments that are due to the selection of different ecotoxicological test data \citep{mark_analysis_1998}. Thereby, Standartox provides the basis for reproducible science and combines information from different sources to simplify the derivation of risk indicators such as Species Sensitivity Distributions (SSD) and Toxic Units (TU), which represent two prominent concepts to assess effects on organisms in ecotoxicology \citep{posthuma_species_2002, kefford_definition_2011, schafer_effects_2011}. Besides aggregating ecotoxicological test results, Standartox provides a concise overview of the tested chemicals, allowing the identification of potential knowledge gaps. Moreover, Standartox could help in reducing the millions of animals used for toxicity testing each year by facilitating access to ecotoxicity data \citep{hartung_chemical_2009}. Standartox comes with two front-ends, a web application: \url{standartox.uni-landau.de} and the R \citep{rcoreteam_language_2017} package \textit{standartox}, providing convenience structures and thereby largely reducing processing time for users.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data Description}
Standartox constitutes a collection of quality checked ecotoxicological test results. It is build on the ECOTOX database \citep{usepa_ecotox_2019} whose data are processed, cleaned and harmonized to retrieve comparable toxicity endpoints. Subsequently, filter and aggregation methods are created to allow for the retrieval of single toxicity equivalents for specific experimental conditions. The ECOTOX database is updated quarterly, providing on average 5228 (2014 - 2019) new toxicity entries. These are included in Standartox with each update.

@@

\subsection{Perspectives}
Novel predictive frameworks incorporating chemical mode of action and species traits emphasize the need for holistic and automated analyses of large-scale ecotoxicological data \citep{malaj_evolutionary_2016, vandenberg_modeling_2019}. Indeed, the increasing amount of data from ecotoxicological tests and experiments that is becoming available has elicited several initiatives to harmonize these data. These initiatives partly aim for overlapping goals, yet have limitations or objectives that distinguish them from Standartox:
\par
\textbf{Comptox}, is a web tool published by the EPA which, similar to Standartox allows for filtering test results, the retrieval of additional chemical information as well as predicted toxicity data \citep{martin_prediction_2001}, such as 48 h \textit{Daphnia magna} LC\textsubscript{50} values. However, toxicity estimations are limited to standard test organisms (e.g. \textit{Daphnia magna}), and the tool lacks the possibility for automated data retrieval \citep{williams_comptox_2017}.
\par
The \textbf{EnviroTox} database which also uses, amongst others the ECOTOX database as an input has recently been published \citep{healthandenvironmentalsciencesinstitutehesi_envirotox_2019, connors_creation_2019}. In contrast to Standartox, EnviroTox  is restricted to selected aquatic organisms (i.e. fish, amphibians, invertebrates and algae) and experimental durations (at least 24 h) and uses a rule-based algorithm to derive single ecotoxicity values. Besides, EnviroTox provides additional information on toxicity endpoints, such as acute or chronic classifications and mode of action assignments. We intentionally omitted such classifications given that the approach to classification may vary with the purpose of the study. The EnviroTox database allows for an aggregation into single toxicity values for individual taxa, whereas Standartox performs this aggregation for individual taxa-chemical combinations. However, the Standartox results for individual taxa-chemical combinations could easily be aggregated across chemicals in a second step to provide a similar aggregation as that performed in EnviroTox. Since Standartox aims to provide toxicity estimations for specific taxa and chemicals we refrain from aggregating further.
\par
The \textbf{Etox} database collects ecotoxicity test information and provides methods to filter those. Like the ECOTOX database, it also doesn't provide methods to perform aggregations of the ecotoxicity data and the database is not accessible in automated way. In contrast to the ECOTOX database it is not possible to download the whole database.
\par
The \textbf{PPDB} provides data only on pesticides, and as mentioned before, it provides single quality controlled values only for commonly used taxa, e.g. \textit{D. magna} or \textit{R. subcapitata}.
\par
In summary, none of the above mentioned initiatives aim for an automated and standardized aggregation method of exposure endpoints for individual chemicals. In addition, they lack the possibility to access the databases through a common high level programming language, such as R.

\begin{table}[H]
    \caption{Overview on databases that provide ecotoxicological data. Abbreviations: ALL: Most important test parameters, including chemical, taxon, duration for filtering ecotoxicological data are incorporated. Web: Accessible via a web application through a graphical user interface. API: Accessible via an application programming interface.}
    \label{tab:database-differences}
    \centering
    \begin{tabular}{ccccc}
    \toprule
    \textbf{Database} & \textbf{Filter} & \textbf{Aggregation, Selection} & \textbf{Access} \\
    \midrule
    Comptox \citep{williams_comptox_2017} & Chemical & no & Web, file \\
    Ecotox \citep{usepa_ecotox_2019} & ALL & no & Web, file \\
    EnviroTox \citep{healthandenvironmentalsciencesinstitutehesi_envirotox_2019} & ALL & chemical, organism & Web \\
    Etox \citep{umweltbundesamt_etox_2019} & ALL & no & Web \\
    Pesticides Properties DataBase (PPDB) \citep{lewis_international_2016} & fixed values & manual selection & Web, file \\
    Standartox & ALL & chemical, organism & API, Web \\
    \bottomrule
\end{tabular}
\end{table}

\par
As outlined above, toxicity estimates from different studies can vary strongly due to a wide range of experimental conditions such as pH, temperature and conductivity \citep{rosenkrantz_influence_2013, li_temperature_2011}. Integrating these conditions into the aggregated estimates would certainly improve toxicity estimates. However, the current implementation of Standartox omits these conditions, because the ECOTOX database only provides sparse records on experimental conditions. The most frequently provided experimental conditions are temperature (77~\%), pH (56~\%), hardness (27~\%), dissolved oxygen (18~\%), Alkalinity (15~\%) and salinity (9~\%). For all other conditions less than 5~\% of data entries are available. A text-mining approach, where a literature reference is associated with ecotoxicity raw data, iterating through the individual publications could potentially increase this number. E.g. \citet{compson_linking_2018} successfully applied text-mining techniques to retrieve species trait data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
An automated processing pipeline downloads the quarterly released ECOTOX database, performs several preparation steps on it and exports a final Standartox data set. This data set is accessible via a web application and an application programming interface (API). An API provides the means for machine communication between a host and a client and thus allows scriptable data queries. To facilitate the API access, the R \citep{rcoreteam_language_2017} package \textit{standartox} is built. All data presented in this paper are derived from the Standartox build based on the ECOTOX release from the 12.12.2019. The code for Standartox is located in two Github repositories \url{https://github.com/andschar/standartox-build} and \url{https://github.com/andschar/standartox}. The former contains code to process the data and to build the web application and the API, the latter contains code to build the R-package. Most of the code is written in R 3.6.1 and associated packages (List: \ref{tab:rpackages}) and SQL for PostgreSQL 9.6.1.

\subsection{Processing}
Standartox downloads the quarterly released ECOTOX database and builds it into a local PostgreSQL database. Subsequently, Structured Query Language (SQL) functions for further processing the data are implemented. In addition lookup tables that enable the conversion of units such as duration and concentration are created. A meta-table providing information such as the release version of the ECOTOX database is added. Then, provided Chemical Abstracts Service Numbers (CAS) and taxonomic names are used to query additional information from publicly available databases on chemicals and organisms, respectively. This includes the Compendium of Pesticide Common Names \citep{wood_compendium_2019}, the Chemical Entities of Biological Interest (ChEBI) database \citep{hastings_chebi_2016}, the Chemical Identifier Resolver (CIR) service \citep{nationalinstitutesofhealthnih_chemical_2019}, the Pubchem database \citep{kim_pubchem_2016}, Eurostat \citep{europeancommission_eurostat_2019} and Wikidata \citep{vrandecic_wikidata_2014} for chemicals and the World Register of Marine Species (WoRMS) \citep{wormseditorialboard_world_2018} and the Global Biodiversity Information Facility (GBIF) \citep{gbif_gbif_2019} (Table~\ref{tab:data-base-additional}) for habitat and spatial distribution of organisms. Given that taxonomic names can be ambiguous, e.g. the genus \textit{Eisenia} can refer to an algae and a worm, we first match the taxa names against specific database identifiers (e.g. WoRMS \citep{wormseditorialboard_world_2020}) and subsequently check their accordance with the underlying ECOTOX data taxonomy. Then we query the actual data by using the identifiers. In a next step the data is added to Standartox to enable filtering for specific chemical roles (e.g. drug, metal, pesticide, personal care product) and classes (e.g. pyrethroid, carbamate) as well as spatial distribution (i.e. continents) and habitat preferences (e.g. freshwater) of individual taxa. Taxa that were not identified to at least genus level are excluded, because relative toxicity comparisons have been shown to be not meaningful for higher taxonomic levels \citep{rainbow_trace_2002, buchwalter_differences_2005, malaj_physiological_2012}. Finally, the Standartox data set is compiled, which includes the harmonisation of data, e.g. through conversion of test concentration and duration units. 1237 distinct concentration units are converted to X harmonized ones, as are the 126 distinct duration units to X. Standartox retains only duration units that can unambiguously be converted to hours. To guarantee appropriate unit conversion and harmonisation, we compared the results of an automated unit conversion to a manual one for each of the distinct concentration and duration units. Furthermore, the units are cleaned, for example through removing additional information in the field such as \textit{food}, \textit{soil}, \textit{ai} that are also coded in other variables and hinder the processing of units. Concentrations that are given as rates such as per day (e.g. mg/kg/day) are multiplied by the days of the test and then converted. Experimental endpoints are restricted to three groups, namely NOEX, LOEX and XX50. Other endpoints, such as Bioconcentration factors, non-half maximal effective concentrations (e.g. IC10, EC25, LD99) or maximum acceptable toxicant concentrations are removed. Along with that, a catalog, listing all distinct entries and value ranges, for categorical and continuous variables, respectively, is created. Finally, we run quality control scripts that check the accuracy of the data. The compiled Standartox data set together with the catalog is exported and accessible via the web application and the API, through the R package.

\begin{figure}[H]
    \centering
    \includestandalone[scale=0.75]{article/tikz/standartox-organigram}
    \caption{Organigram of Standartox. The ECOTOX knowledge base is downloaded quarterly and processed (i.e. query additional information with CAS and taxa names and conversion of concentration and duration units. Subsequently, a Standartox data set is compiled together with filter and aggregation methods. Thus, users can access the Standartox data set and filter and aggregate through a web application and an R-package.}
    \label{fig:stx-organigram}
\end{figure}

\subsection{Application methods}
When accessed, the web application and the API load the compressed serialized Standartox data into memory and allow the user to interact with it. The user can then call the functions \textit{stx\_filter()} and \textit{stx\_aggregate()} that filter and aggregate the data according to specific parameters (Table~\ref{tab:app-parameters}). The interactive web application is built in R using the shiny framework, which runs with the help of a shiny server \citep{chang_shiny_2018} on \url{standartox.uni-landau.de}. The API is built by using the R package \textit{plumber} \citep{trestletechnologyllc_plumber_2018}, which allows for the creation of Representational State Transfer (REST) APIs from R. REST is a software architectural style that defines web service communication rules. The API is reachable via the URL \url{http://139.14.20.252} and port 8000. Four API-endpoints (\textit{/catalog}, \textit{/filter}, \textit{/aggregate} and \textit{/meta}) can be queried (Table~\ref{tab:api-endpoints}). The \textit{/catalog} API-endpoint returns a JavaScript Object Notation (JSON) file containing a catalog of possible filter parameters to choose from. The \textit{/filter} returns the filtered Standartox table as a compressed serialized binary file created by the R \textit{fst} package \citep{klik_fst_2019}, to reduce size and allow for fast user queries. The \textit{/aggregate} API-endpoint returns an R function allowing the client to aggregate the filtered data locally. Lastly, the \textit{/meta} API-endpoint returns a JSON file with meta information, such as the timestamp of the request and the used Standartox version. The API is designed to be used with the R-package \textit{standartox} and therefore uses serialization methods specific to R (rds() from the R package \textit{base} and fst() from the package \textit{fst}). To facilitate the API usage the R-package \textit{standartox} is created.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{User Notes}
Users can access Standartox either via the web application \url{standartox.uni-landau.de} or via the R-package \textit{standartox}. By accessing the web application, users can filter and download the resulting data sets as a comma-separated values (csv) file. Users of the R-package can directly load the data within R. The R-package provides the two functions \textit{stx\_catalog()} and \textit{stx\_query()}. The first command queries a catalog of possible Standartox parameters into an R list object. The latter allows users to set the Standartox filter parameters and to fetch the actual data. It returns an R list of three tables (i.e. R \textit{data.frame}s) containing the filtered data set, the aggregated data set and a table with the meta information retrieved from the API endpoints. A short R-code example is given below (Listing \ref{lis:standartox-example}) and a detailed description on the usage of the R package is provided on its Github page: \url{https://github.com/andschar/standartox}.

\begin{lstlisting}[
    caption = {Sample code to access the Standartox database through the API and the the R-package \textit{standartox}. \textit{stx\_catalog()} returns a catalog of possible filter and aggregation parameters. \textit{stx\_query()} returns the Standartox object, a list of the filtered and the aggregated data as well as a meta data entry. Example for XX50 tests on the compound glyphosate (CAS: 1071-83-6) and the taxon \textit{Oncorhynchus} lasting 24 h to 120 h.},
    label = {lis:standartox-example}]
# install
install.packages('remotes')
remotes::install_github('andschar/standartox')
# retrieve catalog    
require(standartox)
catal = stx_catalog()
# retrieve data
l = stx_query(cas = '1071-83-6',
                endpoint = 'XX50',
                taxa = 'Oncorhynchus',
                duration = c(24, 120))
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
% This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
Due to the steady incorporation of new ecotoxicity data, the aggregated values produced by Standartox can be subject to change with future updates. We regard this as an advantage rather than a drawback because other published works that aim in a similar direction often constitute a singular effort or require manual work for each update. Standartox, in contrast automates the update process, yet still provides access to its older versions, assuring reproducibility and version control. In comparison to rule-based approaches for the derivation of single ecotoxicity values, Standartox has the advantage to be free from the subjectivity of a set of human-induced rules. Above all, Standartox provides quick access through its design to be queried via the R language. Due to an increased amount of available ecotoxicological test data, it becomes fundamental to provide and distribute ecotoxicity information in adequate formats, both easily accessible for humans and easily processable for machines. Standartox meets these requirements and puts its focus on the aggregation of toxicity data, thereby adding a piece to the puzzle of modern ecotoxicological data analyses.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{conceptualization and methods, A.S., R.B.S. and V.C.S.; software development and validation, A.S.; data curation and analysis, A.S.; writing--original draft preparation, A.S.; writing--review and editing, A.S., V.C.S. and R.B.S.; supervision, R.B.S.}
% \authorcontributions{conceptualization, A.S., R.B.S. and V.C.S.; methodology, A.S.; software, A.S.; validation, A.S., Y.Y. and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing--original draft preparation, X.X.; writing--review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y.''}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{The authors want to thank Eduard Szöcs whose blog post (https://edild.github.io/localecotox/) on how to build a local version of the EPA ECOTOX database was the starting point for this project.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{The authors declare no conflict of interest.} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{The following abbreviations are used in this manuscript:\\
\noindent 
\begin{tabular}{@{}ll}
API & Application programming interface \\
CAS & Chemical abstracts service registry Number \\
ChEBI & Chemical Entities of Biological Interest database \\
CIR & Chemical Identifier Resolver service \\
E/LC/D\textsubscript{50} &  Half maximal effective/lethal concentration/dose \\
ECOTOX & US EPA Ecotox Knowledgebase \\
GBIF & Global Biodiversity Information Facility \\
JSON & JavaScript Object Notation file format \\
LOEC/L & Lowest observed effect levels/concentration \\
NOEC/L & No observed effect levels/concentration \\
PPDB & Pesticides Properties DataBase \\
QSAR & Quantitative structure–activity relationship \\
REST & Representational state transfer software architectural style \\
SSD & Species sensitivity distribution \\
TU & Toxic unit \\
WoRMS & World Register of Marine Species \\
\end{tabular}
}
